Density-dependent selection and the limits of relative fitness

### 2017-12-19 manuscript.tex
l

In uncrowded populations, relative fitness simply represents differences in intrinsic population growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, it seems likely that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}). Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

Counter to the assumption of constant $N$, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

In light of these difficulties, the relative fitness description has been justified in broadly three different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology). The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp. 468]{barton_2007} \citep{prout_1980}. Obviously this  does not address the problem that $s$ can, in reality, depend on density. Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

@@

Third, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.

Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}. Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

@@

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.

>>>>

In age-structured, sexual populations, selection cannot in general be represented in terms of a single type-specific fitness parameter \citep[Chap. 4]{charlesworth_1994}. Relative fitness models can then be justified as a linear approximation in which selection is approximately constant \cite[pp. 277]{ewens_2004} \citep[Chap. 4]{charlesworth_1994}. Provided that selection is sufficiently weak and stable over time, this linear approximation is not a major restriction on the use of relative fitness. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.

Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``linear approximation'' restriction or the poor integration with population ecology \citep{mallet_2012}. Thus, it is important that we understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

Our aim here is to explore in more depth where the population ecological limits of relative fitness lie. To do so, we set aside the complexities of age structure and mating, since we know that these can cause relative fitness to break down. In this simplest case of asexual haploids with no age structure, the primary threat to the relative fitness description of selection is the dependence of selection on population density.

To see this, first consider an uncrowded population. Relative fitness then simply represents differences in intrinsic population growth rate. In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$. Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$. In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp. 26]{crow_1970}. If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}). Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

However, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}. The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b), even in the absence of age-structure and mating.

@@

In light of these difficulties, the relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology). The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp. 468]{barton_2007} \citep{prout_1980}. Obviously this does not address the problem that $s$ can, in reality, depend on density. Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.

@@

@@

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group'' \cite[pp. 54]{charlesworth_1994}). In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.
